ANNO: The Anthropological Notation Ontology

### 2023-04-04 paper.tex
Exchange 3-ontology text with Alex' version. Add preliminarly ANNO-core image.

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
\usepackage{todonotes}
\usepackage{cleveref}
%\usepackage{dcolumn}

@@

\section{Developing ANNO}
The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
The \emph{domain ontology} describes the domain of skeletal anatomy.
The \emph{task ontology}
The General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} and integrates the domain and task ontologies.

% todo: do we have an ontology engineering methodology?
% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
% TODO: ontology language, expressivity
% TODO: reuse of ontological entities
% TODO: clear validation mechanism
% TODO: top-level ontology?
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}
The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.\todo{explain}

@@

\begin{figure}[h]
\includegraphics[width=0.3\textwidth]{img/modules.pdf}
\caption{The six modules of ANNO.}\label{fig:modules}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{img/axisplane.pdf}
\caption{Anatomical axes and planes.}\label{fig:axisplane}
\end{figure}

@@

TODO: Modellierung der Anthropologie-Domäne (u.a. Motivation, related work, was ist neu).
TODO: Dann könnte man Design-Entscheidungen beschreiben (wie und warum wurden bestimmte Sachverhalte so in der Ontologie modelliert).
\section{Publication}
ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
TODO: Bereitstellung der Ontologie für die Community (OLS etc.).

>>>>

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
\usepackage{todonotes}
\usepackage{cleveref}
\usepackage{csquotes}
%\usepackage{dcolumn}

@@

\section{Developing ANNO}
%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
%First, \emph{task ontology} is created as a conceptual model of \aw{}.
%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
\todo{figure out how to best separate ontology and software}
The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.

In our case, the ANNO-core (\cref{fig:core}) functions as a TO.
It contains the general anatomical entities (\enquote{Anatomical structure} and \enquote{Landmark}),
\enquote{Spatial relator} (for describing measurements between landmarks, e.g., distances or angles),
\enquote{Phenotype} (to model properties that can be derived from the measurements) and \enquote{Anatomical spatial entity} (to describe anatomical directions, axes and planes).

The subclasses of the ANNO-Core classes build a DO, the ANNO-domain.
These include concepts representing concrete bones (e.g., mandible) and teeth (e.g., left mandibular central secondary incisor tooth),
specific landmarks on them (e.g., mentale dexter and mentale sinister),
concrete measurements/relators (e.g., the distance between mentale dexter and mentale sinister),
individual phenotypes (e.g., sex or height) and defined anatomical directions (e.g., cranial or basal),
axes (e.g., longitudinal or sagittal axis) and planes (e.g., frontal or sagittal plane).

For the foundation of the TO we used the \emph{General Formal Ontology} (GFO)~\citep{gfo} as TLO.
In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Property} as well as the spatial entities of GFO (Fig. …).
One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) of the TO (ANNO-core),
whereas the entities of the corresponding DO (ANNO-domain) are processed dynamically.
\aw{} uses the ANNO-core as an interface to access the ANNO-domain entities (e.g., by listing all subclasses of a specific ANNO-core class).

% todo: do we have an ontology engineering methodology?
% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
% TODO: ontology language, expressivity
% TODO: reuse of ontological entities
% TODO: clear validation mechanism
% TODO: top-level ontology?
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}
The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
\todo{können wir es auf OWL 2 DL eingrenzen?}
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.\todo{explain}

@@

\begin{figure}[h]
\includegraphics[width=0.6\textwidth]{img/core.pdf}
\caption{ANNO-core ontology}\label{fig:core}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.3\textwidth]{img/modules.pdf}
\caption{The six modules of ANNO}\label{fig:modules}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{img/axisplane.pdf}
\caption{Anatomical axes and planes}\label{fig:axisplane}
\end{figure}

@@

TODO: Modellierung der Anthropologie-Domäne (u.a. Motivation, related work, was ist neu).
TODO: Dann könnte man Design-Entscheidungen beschreiben (wie und warum wurden bestimmte Sachverhalte so in der Ontologie modelliert).
\section{Publication}
ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
TODO: Bereitstellung der Ontologie für die Community (OLS etc.).
